---
title: "Session 2: Bayesian Mixed-Effects Models with brms"
subtitle: "Ex-Gaussian RTs, Bernoulli Accuracy, and Shifted Lognormal Eye Movements"
author: "Dr. Bernhard Angele"
format:
  revealjs:
    incremental: true
    theme: [simple, custom.scss]
    logo: images/logo.png
    slide-number: true
    toc: true
bibliography: references.bib
csl: apa.csl
---

## Session Roadmap (3h)

- **Part 1 (45 min):** The Bayesian multilevel framework, priors, and response families
- **Part 2 (45 min):** Modeling response times with the ex-Gaussian distribution in `brms`
- **Part 3 (45 min):** Modeling accuracy (Bernoulli) & complex random effect structures
- **Part 4 (45 min):** Modeling eye movements (Shifted Lognormal), diagnostics, and LOO model comparison

## Section 1: The Bayesian Multilevel Framework {.dark-bg background-color="#1F497D"}

## Why Bayesian Multilevel Models?

Frequentist models (`lme4`) are highly efficient but suffer from critical limitations:

1. **Singular Fit Failures:** Complex maximal random-effect structures often fail to converge or result in zero-variance estimates.
2. **Strict Likelihood Assumptions:** RTs and eye-tracking durations are highly skewed, forcing arbitrary log-transformations.
3. **No Direct Probability:** Confidence intervals and $p$-values are frequently misinterpreted; they do not state the probability of an effect being true.

## The Bayesian Solution

::: {.accent-box}
### Three Bayesian Columns
* **Priors:** Regularize parameters and prevent singular fits through shrinkage.
* **Flexible Families:** Model skewed distributions (ex-Gaussian, Shifted Lognormal) directly on the raw scale.
* **Posterior Distributions:** Obtain direct probability statements regarding our hypotheses.
:::

## Prior Specification

Bayesian inference combines our **Priors** (prior beliefs about parameter scales) and **Likelihood** (observed data) to calculate the **Posterior** distribution:

* **Weakly Informative/Regularizing Priors:** Restrict parameters to physically/physiologically plausible bounds (e.g., response times cannot be negative or 10 seconds).
* **Shrinkage:** Pulls extreme, noisy random effect estimates towards the group mean, preventing overfitting.

## Section 2: Modeling RTs with the ex-Gaussian Distribution {.dark-bg background-color="#1F497D"}

## What is the ex-Gaussian Distribution?

Human reaction times are rarely symmetrical. They are heavily right-skewed. The **ex-Gaussian** distribution represents this shape by combining:

1. A **Gaussian (normal) distribution** representing the cognitive core ($\mu$ and $\sigma$).
2. An **exponential distribution** representing the right-tail skew ($\beta$).

$$\text{RT} \sim \text{Normal}(\mu, \sigma) + \text{Exponential}(\beta)$$

Rather than log-transforming response times, we can model raw `rt` directly and test how our conditions affect both the **core latency** ($\mu$) and the **decision tail** ($\beta$).

## Bayesian RT: Priors & ex-Gaussian setup

We will define regularizing priors for both the Gaussian core ($\mu$) and the exponential tail ($\beta$) of our reaction times before fitting the model in `brms`:

```{r}
#| eval: false
library(brms)
library(tidyverse)

# Define regularizing priors on the log/raw scale
priors_exg <- c(
  prior(normal(0.6, 0.2), class = "Intercept"),       # RT intercept (600ms)
  prior(normal(0, 0.1), class = "b"),                 # Fixed effects on RT
  prior(exponential(2), class = "sd"),                # Subject/Item variation
  prior(lkj(2), class = "cor"),                       # Random correlation matrix
  
  # Priors for the exponential parameter beta
  prior(normal(-1.5, 0.5), class = "Intercept", dpar = "beta"), 
  prior(normal(0, 0.3), class = "b", dpar = "beta")
)
```

## Fitting ex-Gaussian in brms (Command)

```{r}
#| eval: false
# Fit ex-Gaussian model on Flanker Exp 1 correct RTs
blmm_rt <- brm(
  formula = bf(
    rt ~ Condition * StimulusType + (Condition * StimulusType | subject) + (Condition | item),
    beta ~ Condition * StimulusType + (Condition * StimulusType | subject) + (Condition | item)
  ),
  data = flanker_rt_data,
  family = exgaussian(),
  prior = priors_exg,
  chains = 4, iter = 4000, warmup = 2000,
  control = list(adapt_delta = 0.95),
  cores = 4, backend = "cmdstanr"
)
```

## Section 3: Modeling Accuracy (Bernoulli) {.dark-bg background-color="#1F497D"}

## Bernoulli GLMM for Binary Accuracy

For single-trial binary accuracy data (`corr` = 0 or 1), the binomial model is simplified to a **Bernoulli distribution** with a logit link function:

$$\text{corr} \sim \text{Bernoulli}(p)$$
$$\text{logit}(p) = \log\left(\frac{p}{1-p}\right) = X\beta + Z u$$

This models the probability $p$ of a correct response directly on the log-odds scale, utilizing a maximal random effects structure:

## Fitting Bernoulli in brms (Command)

```{r}
#| eval: false
# Fit Bernoulli multi-level model on all Flanker Exp 1 trials
blmm_acc <- brm(
  formula = corr ~ Condition * StimulusType + 
    (Condition * StimulusType | subject) + 
    (Condition | item),
  data = flanker_clean,
  family = bernoulli(link = "logit"),
  prior = prior(normal(0, 1), class = "b"),
  chains = 4, iter = 4000, warmup = 2000,
  cores = 4, backend = "cmdstanr"
)
```

## Section 4: Modeling Eye Movements & Model Comparison {.dark-bg background-color="#1F497D"}

## Shifted Lognormal for Eye-Tracking

Like manual reaction times, eye-tracking duration measures (e.g. Total Viewing Time `tvt`) are highly skewed.
However, eye-tracking fixations have a **physiological minimum duration** (typically around 50–80 ms) below which a fixation cannot be registered.

::: {.accent-box}
### The Shifted Lognormal Distribution
The **Shifted Lognormal** distribution models the outcome on a log scale but adds a shift parameter ($ndt$: non-decision time) representing this absolute physiological floor:
$$\text{tvt} \sim \text{Lognormal}(\mu, \sigma) + ndt$$
:::

## Fitting Eye-Tracking LMM in brms (Command)

```{r}
#| eval: false
# Fit Shifted Lognormal LMM on Gaze-Contingent target word Total Viewing Time
blmm_tvt <- brm(
  formula = tvt ~ preview * number * target_gender + (1 | subject) + (1 | sentence_nr),
  data = gaze_target %>% filter(tvt > 50 & tvt < 4000),
  family = shifted_lognormal(),
  prior = c(
    prior(normal(5.5, 0.5), class = "Intercept"),
    prior(normal(0, 0.2), class = "b"),
    prior(exponential(1), class = "sd")
  ),
  chains = 4, iter = 4000, warmup = 2000,
  cores = 4, backend = "cmdstanr"
)
```

## Bayesian MCMC Diagnostics

Before interpreting posterior parameters, we MUST verify the sampling quality:

1. **R-hat ($\hat{R}$):** Should be exactly $< 1.01$. Verifies that all chains converged onto the same posterior target.
2. **Effective Sample Size (ESS):** Bulk and Tail ESS should be $> 1000$ to ensure stable posterior statistics.
3. **Divergent Transitions:** Stan will warn if MCMC chains encounter numerical instability.
   - *Fix:* Increase `control = list(adapt_delta = 0.98)`.

## Posterior Predictive Checks

A powerful feature of Bayesian models is simulating new datasets from the posterior and overlaying them on the actual data to see if the model successfully captured the shape, skewness, and boundaries:

* **Visual Alignment:** We expect the simulated density curves to line up closely with the raw empirical data density.
* **Warning Signs:** Mismatches in the peak height or tails indicate that the distribution family is poorly specified.

## Generating Checks in R (Command)

```{r}
#| eval: false
# Generate posterior predictive checks
pp_check(blmm_rt, ndraws = 100)
pp_check(blmm_tvt, ndraws = 100)
```

## Model Comparison via LOO-CV

To compare competing models (e.g. whether a log-normal distribution or a shifted-lognormal distribution fits our eye-tracking data better), we use **Leave-One-Out Cross-Validation (LOO-CV)**:

* **Expected Log Predictive Density (`elpd`):** Quantifies expected out-of-sample predictive accuracy.
* **Comparison Matrix:** Models are ordered by predictive performance. Competing models are compared to the best model using standard error differences.

## Computing LOO in R (Command)

```{r}
#| eval: false
# Compute LOO criteria
loo_model1 <- loo(blmm_tvt)
loo_model2 <- loo(blmm_tvt_lognormal)

# Compare expected log pointwise predictive density (elpd)
loo_compare(loo_model1, loo_model2)
```

## Posterior Inferences & Hypothesis Testing

In the Bayesian framework, we do not need arbitrary $p$-values. We can draw posterior samples and directly compute the probability of our experimental hypotheses:

* **Directional Probability:** e.g., what is the probability that our concordant flanker effect is less than zero ($P(\beta < 0)$)?
* **Credible Intervals:** Express uncertainty directly (e.g., "There is a $95\%$ probability that the true effect lies between $-35\text{ms}$ and $-5\text{ms}$").

## Querying Posteriors in R (Command)

```{r}
#| eval: false
library(tidybayes)

# Get posterior samples for the flanker condition contrast
post_samples <- blmm_rt %>%
  spread_draws(b_ConditionCON_vs_AGR)

# What is the probability that Concordant flankers are faster than Control?
# (i.e. that the difference is negative)
prob_faster <- mean(post_samples$b_ConditionCON_vs_AGR < 0)
cat("P(Concordant < Control) =", prob_faster * 100, "%\n")
```

## Session 2 Summary

1. **Bayesian Advantages:** Regularizing priors prevent singular fits, while flexible likelihood families model real-world distributions directly.
2. **ex-Gaussian RTs:** Models core cognitive speed ($\mu$) and exponential skew tail ($\beta$) simultaneously.
3. **Bernoulli Accuracy:** Fits single-trial binary accuracy without aggregation.
4. **Shifted Lognormal Eye Movements:** Respects the physiological minimum duration limit of fixations.
5. **MCMC Diagnostics & Verification:** Always verify $\hat{R} < 1.01$ and run `pp_check` to validate predictive accuracy.

## References

::: {#refs}
:::

