---
title: "Session 1: Frequentist Mixed-Effects Models with lme4"
subtitle: "Grammatical Gender Flanker & Gaze-Contingent Boundary Paradigms"
author: "Dr. Bernhard Angele"
format:
  revealjs:
    incremental: true
    theme: [simple, custom.scss]
    logo: images/logo.png
    slide-number: true
    toc: false
bibliography: references.bib
csl: apa.csl
---

## Workshop plan

- **Session 1 (today)**
  - **Part 1** Introducing the workshop data sets
  - **Part 2:** Data preprocessing, exclusions, and custom contrast coding
  - **Part 3:** Linear mixed-effects modeling of response times (Gaussian) using `lme4`

- **Session 2 (tomorrow)**
  - **Part 1** The Bayesian multilevel framework, priors, and response families
  - **Part 2:** Modeling response times with the Gaussian distribution in `brms`
  - **Part 3:** Using the ex-Gaussian distribution in `brms`
  - **Part 4:** Model diagnostics, posterior predictive checks, interpretation

## Part 1: Introduction to the Workshop Datasets {.dark-bg background-color="#C2002F"}

## Parafoveal Processing of Articles in Spanish

Our workshop utilizes a dataset we recently published, exploring how readers process grammatical gender in Spanish:

Experiment 1 from @serrano-carot2026lcn used a **word-flanker paradigm** to investigate how **parafoveal syntactic processing** occurs during single-word reading. Participants performed different tasks (Exp 1: grammatical gender categorization, Exp 2: lexical decision, Exp 3: semantic categorization) while being presented with target nouns flanked by either concordant, discordant, or neutral articles. In this session, we will focus on the data from Experiment 1, the grammatical gender categorization task.

## Why Study Spanish Articles?

Unlike English, which features a single gender-neutral definite article (*the*), Spanish has a morphosyntactically rich article system:

* **Gender Agreement:** *el* (masculine), *la* (feminine)
* **Number Agreement:** *los* (masculine plural), *las* (feminine plural)

This rich variation makes Spanish an ideal linguistic test-case for investigating **parafoveal syntactic processing** [@angele2013].

## The Grammatical Gender Agreement Effect

* **Linguistic Rule:** Spanish articles must agree in gender and number with the noun they modify [@corbett1991].
* **Violations:** Agreements like **\*el mesa** (masculine article + feminine noun) cause severe cognitive disruption:
  - Delayed response times in foveal categorization and lexical decisions [@barber2005].
  - Distinct neurophysiological ERP signatures, such as the **P600** component [@alemánbañón2016].
* **Core Research Question:** Are these syntactically complex agreement rules processed *parafoveally* before fixating the noun?


## Dataset: Grammatical Gender Flanker (Exp 1)

- **Source Study:** @serrano-carot2026lcn
- **Task:** Noun grammatical gender decision task. Participants decide if a target noun (e.g., *mesa*, feminine; *sol*, masculine) is masculine or feminine.
- **Flanker Design:** The target noun is flanked by articles that are either:
  - **Concordant (`AGR`):** Flankers agree in gender (e.g., *la mesa la*).
  - **Discordant (`DIS`):** Flankers mismatch in gender (e.g., *el mesa el*).
  - **Control (`CON`):** Non-gender characters flanking the noun.
- **Key Outcomes:** Response Time (`rt` in seconds) and Response Accuracy (`corr`: 1 = correct, 0 = incorrect).

## Flanker Paradigm: Conditions

:::: {.columns}
::: {.column width="33%"}
**Control (`CON`):**
```{=html}
<div class="flanker-container">
  <div class="flanker-stage cross">+</div>
  <div class="flanker-stage stimulus">mesa</div>
  <div class="flanker-stage target-only">mesa</div>
  <div class="flanker-stage feedback-correct">✅</div>
  <div class="flanker-stage feedback-incorrect">❌</div>
</div>
```
No flankers (target presented on its own).
:::

::: {.column width="33%"}
**Concordant (`AGR`):**
```{=html}
<div class="flanker-container">
  <div class="flanker-stage cross">+</div>
  <div class="flanker-stage stimulus">la&nbsp;&nbsp;mesa&nbsp;&nbsp;la</div>
  <div class="flanker-stage target-only">mesa</div>
  <div class="flanker-stage feedback-correct">✅</div>
  <div class="flanker-stage feedback-incorrect">❌</div>
</div>
```
Flankers agree in grammatical gender.
:::

::: {.column width="33%"}
**Discordant (`DIS`):**
```{=html}
<div class="flanker-container">
  <div class="flanker-stage cross">+</div>
  <div class="flanker-stage stimulus">el&nbsp;&nbsp;mesa&nbsp;&nbsp;el</div>
  <div class="flanker-stage target-only">mesa</div>
  <div class="flanker-stage feedback-correct">✅</div>
  <div class="flanker-stage feedback-incorrect">❌</div>
</div>
```
Flankers violate gender agreement.
:::
::::

---

## Section 2: Data Preprocessing & Custom Contrasts {.dark-bg background-color="#C2002F"}

## Introduction to the Tidyverse

We will use the **tidyverse** ecosystem for all data preprocessing, cleaning, and visualization.

* **What is it?** An opinionated collection of modern R packages designed for data science.
* **Why use it?** 
  - Packages share a common philosophy, design token, and grammar.
  - Expressive syntax (`%>%` pipe operator) that reads like a sequence of actions.
  - Highly efficient for transforming messy experimental logs into analysis-ready data frames.
* **Core Packages in Our Workflow:**
  - `readr`: Fast, friendly reading of CSV files.
  - `dplyr`: Data manipulation verbs like `filter()`, `mutate()`, and `select()`.
  - `ggplot2`: Data visualization based on the grammar of graphics.

## Flanker Experiment Dataset Structure

The raw dataset `data/exp1_data_for_analysis.csv` contains trial-level experimental logs:

::: {style="font-size: 0.65em;"}

| Column | Type | Description | Example |
|---|---|---|---|
| `subject` | integer | Unique subject index | `1` |
| `Target` | character | Target noun presented on the screen | `"mesa"` |
| `Condition` | character | Flanker condition (`CON`, `AGR`, `DIS`) | `"AGR"` |
| `StimulusType` | character | Target grammatical gender (`MAS` or `FEM`) | `"FEM"` |
| `rt` | numeric | Reaction time in seconds | `0.3273` |
| `corr` | integer | Accuracy (`1` = correct, `0` = incorrect, `-1` = timeout) | `1` |
| `Flanker` | character | Specific article flanking the noun | `"la"` |

: Key columns of the flanker experiment dataset {.striped .hover}

:::

## Step 1: Preprocessing the Flanker Data

Before fitting models, we must apply standard exclusions to filter out noise:

1. **Timeout Exclusions:** Exclude trials where participants timed out (`corr == -1`).
2. **Physiological Plausibility Window:** Retain manual reaction times strictly between $250\text{ ms}$ and $1800\text{ ms}$ ($0.25\text{s} - 1.8\text{s}$).
3. **Counterbalancing & Subject Factors:** Cast key variables (Subject, Target Word, Condition, Gender) as clean factors in R.

## Loading & Cleaning in R

```{r}
#| eval: true
#| echo: false
library(tidyverse)
library(lme4)
library(MASS)
library(here)

# Load Flanker Exp 1 data
flanker_raw <- read_csv(here("data/exp1_data_for_analysis.csv"))

# Exclude timeouts and extreme response times
flanker_clean <- flanker_raw %>%
  filter(
    corr != -1,
    rt >= 0.25 & rt <= 1.8
  ) %>%
  mutate(
    Condition = factor(Condition, levels = c("CON", "AGR", "DIS")),
    StimulusType = factor(StimulusType, levels = c("MAS", "FEM")),
    subject = factor(PROLIFIC_PID),
    item = factor(Target)
  )
```

## Step 2: Contrast Coding

We have three levels in `Condition` (`CON`, `AGR`, `DIS`). We need to think about what comparisons are the most interesting for us theoretically. In this case, we care mainly about the impact of the dissimilar flankers (DIS) compared to the other two conditions. 
We also wonder whether the concorndant flanker has a facilitation effect compared to the control (CON).

- Why don't we just use the default treatment contrasts? They would compare each level to a baseline (e.g., `CON`). This would not be terrible, but we would lose out on the direct comparison between `AGR` and `DIS`, which is where we would expect the largest difference if the flankers are indeed processed parafoveally. In this case, we decide to use **custom contrasts** to examine the effect of the discordant flanker compared to the concordant and control conditions (i.e., we test the hypothesis that $\frac{AGR+CON}{2} - DIS = 0$) and and the effect of the concordant flanker compared to the control ($AGR - CON = 0$).

## Custom Contrast Coding

We define a **custom contrast matrix** to test:

1. **CON vs AGR (Facilitation):** Does a concordant flanker help compared to a neutral control? $1 \cdot AGR - 1 \cdot CON + 0 \cdot DIS = 0$
2. **CON & AGR vs DIS (Interference):** Does a discordant flanker slow down processing compared to the other conditions? $0.5 \cdot AGR + 0.5 \cdot CON - 1 \cdot DIS = 0$

- We can specify this contrast matrix in the differences format: 

| Condition | CON vs AGR | (CON & AGR) vs DIS |
|---|---|---|
| CON | -1 | 0.5 |
| AGR | 1 | 0.5 | 
| DIS | 0 | -1 |

## Stimulus Type

* We might also want to test whether there is a difference in processing between masculine and feminine nouns. We can use **sum contrasts** for this factor, which will compare each level to the grand mean (i.e., `MAS` vs `FEM`). For this, we can use the R command `contr.sum(2)` and don't have to worry about the general inverse. The contrast will automatically be correctly coded as `MAS = 0.5` and `FEM = -0.5`.

## Orthogonality of Contrasts

* Why did we use the specific custom contrasts for `Condition` and sum contrasts for `StimulusType`? 
    * This ensures that all the contrasts are **orthogonal** (i.e., uncorrelated). When we use correlated predictors in a regression model, this can reduce both the interpretability and statistical power of our model.
* If the model includes interactions, non-orthogonal contrasts can lead to high degrees of multicollinearity, which can inflate standard errors and make it difficult to detect significant effects. By using orthogonal contrasts, we can ensure that each contrast tests a unique aspect of the data without overlap.

## Contrast Setup in R

```{r}
#| eval: true
# Define custom hypothesis contrast matrix
contr_matrix <- matrix(c(
   1, -1,  0,    # Contrast 1: CON vs AGR
   0.5, 0.5, -1  # Contrast 2: (CON + AGR) vs DIS
), nrow = 3, dimnames = list(c("CON", "AGR", "DIS"), c("CON_vs_AGR", "CON_AGR_vs_DIS")))

# Apply the generalized inverse to map contrasts onto regression coefficients
contr_matrix_inv <- zapsmall(t(ginv(contr_matrix))) %>%
  matrix(nrow = 3, dimnames = list(c("CON", "AGR", "DIS"), c("CON_vs_AGR", "CON_AGR_vs_DIS")))

# Assign to factor Condition
contrasts(flanker_clean$Condition) <- contr_matrix_inv
contrasts(flanker_clean$StimulusType) <- contr.sum(2) # Sum code Gender (-1, 1)

# Filter correct trials for LMM
flanker_rt_data <- flanker_clean %>% filter(corr == 1)
```

## Why the generalized inverse?

- There are two ways to specify contrasts in R: the **differences format** (where each row corresponds to a condition and columns correspond to contrasts) and the **regression format** (where each column corresponds to a regression coefficient). The `contrasts()` function in R expects the regression format, so we use the generalized inverse (`ginv()`) from the `MASS` package to convert our differences contrast matrix into the appropriate regression format.

- We could specify the contrasts directly in regression format, but using the differences format is often more intuitive when designing contrasts based on theoretical hypotheses.

- `t()` transposes the matrix (rows to columns and vice versa), and `zapsmall()` is used to clean up any very small numerical values that may arise from the calculation (rounding errors).

- For a detailed explanation of contrast coding and the use of the generalized inverse, see @schad2020capitalize.

## Section 3: Mixed-Effects Modeling: RT & Accuracy {.dark-bg background-color="#C2002F"}

## Why do we need mixed-effects models here?

Independence assumptions of traditional ANOVA are violated in our experimental design:

- **Repeated Measures:** Each participant contributes multiple trials. All measurements from the same participant share variance and are not independent.

- **Crossed Items:** Target words are repeated across subjects. Measurements from the same item may also share variance and not be independent.

- The **Language-as-fixed-effect fallacy:** ANOVAs can only model one source of random variability (e.g., subjects or items), but not both simultaneously [@clark1973language]. This can lead to inflated Type I error rates and incorrect inferences about our fixed effects [@raaijmakers1999deal].

## Fit the full random effects structure
* **Important note:** Unlike repeated measures ANOVAS, linear mixed models allow us to specify which random effects of subject or item we want to include. We can include:
  - **Random Intercepts:** Account for baseline differences in response times across subjects and items (the "main" effect of subject/item).
  - **Random Slopes:** Account for how the effect of our experimental conditions (e.g., Condition, StimulusType) may vary across subjects and items. For example, some subjects may show a larger flanker effect than others, and some items may be more affected by the flanker manipulation than others. This is the "interaction" of subject/item with our fixed effects.

* The equivalent of an F1 ANOVA is not an LMM with only random subject intercepts. It is an LMM with random intercepts and slopes for subjects. If we do not include random slopes, but there are random effects in the data, these may be confounded with fixed effects, inflating Type I error rates [@barr2013].

## Problems with maximal random effects structures in *lme4*

* **Convergence Issues:** If there are no systematic random effects in the data, the model may fail to converge instead of estimating a variance of zero. However, a failure to converge does not necessarily mean that there is no random effect in the data. 
* **Power Issues:** Even if the model converges, it may require a substantially larger sample size to detect fixed effects while also estimating a complex random effects structure. This can lead to Type II errors [failing to detect a true effect, @matuschek2017balancing].

## Fitting LMM in R (Command)

We fit a linear mixed-effects model on correct trials, modeling reaction times (`rt`) as a Gaussian outcome using the `lme4` package.

- **Question**: Why do we include StimulusType as a random slope for subjects but not items? 
  - Because each item only has one gender!

```{r}
#| eval: false

m_rt <- lmer(
  rt ~ Condition * StimulusType + 
    (1 + Condition * StimulusType | subject) + 
    (1 + Condition | item),
  data = flanker_rt_data
  )
```

## This does not converge!

- What to do?
- Simplify the random effects structure. We can start by removing correlations between random effects using the double-pipe syntax `||` in `lme4`. This will allow the model to estimate variances for each random effect without estimating correlations between them. Since the correlations will often be close to zero, this can help with convergence without losing much information about the random effects structure.

```{r}
#| eval: false
m_rt <- lmer(
  rt ~ Condition * StimulusType + 
    (1 + Condition * StimulusType || subject) + 
    (1 + Condition || item),
  data = flanker_rt_data
)
```

## Still does not converge!

- Now we can systematically remove random slopes that we suspect have low variance (e.g., target item slopes) while retaining slopes directly tied to our primary hypothesis (e.g., Condition slopes for subjects). This is a common approach to achieve convergence while still modeling the most important sources of random variability in the data.

```{r}
#| eval: true
#| cache: true
m_rt <- lmer(
  rt ~ Condition * StimulusType + 
    (1 + Condition || subject) + 
    (1 | item),
  data = flanker_rt_data
)
``` 
- Note that we cannot use `||` in the `(1 | item)` term since there is only one random effect here.

## This converges!

We inspect the results using `summary(m_rt)` to retrieve the parameters and statistics:

```{r}
summary(m_rt)
```

## Interpreting the model output:
- **Fixed Effects Estimates ($b$):** Coefficients representing the influence of Condition and Gender (StimulusType) on response times.
- **Random Effects Variances:** Subject-level and Item-level standard deviations.
- **Optimizer Checks:** Verify model convergence and inspect fit statistics.
- **Correlation matrix:** Check for correlations between fixed effects. High correlations (|r| > 0.7) can indicate multicollinearity.

## Interpreting the results:

- **Main Effect of Condition:** 
  - **Contrast 1:** Significant negative $b$ for `CON_vs_AGR` indicates that the concordant flanker condition leads to faster response times compared to the control, consistent with a facilitation account.
  - **Contrast 2:** Significant negative $b$ for `CON_AGR_vs_DIS` indicates that the discordant flanker condition leads to slower response times compared to the other conditions, consistent with an interference account.
- **Main Effect of StimulusType:** Significant positive $b$ for `StimulusType` indicates that masculine nouns (coded as 0.5) are processed slower than feminine nouns (coded as -0.5).
- **Interaction:** No significant interaction terms suggests that the flanker effects do not differ between masculine and feminine nouns.

## Issues with this model:
 - Normality of residuals may be violated due to the right-skewed distribution of response times.
 - Let's have a look at the distribution of the residuals:
```{r}
#| eval: true
# ggplot of residual distribution
residuals <- resid(m_rt)
ggplot(data.frame(residuals), aes(x = residuals)) +
  geom_histogram(binwidth = 0.05, fill = "steelblue", color = "black") +
  labs(title = "Distribution of Residuals", x = "Residuals", y = "Frequency") +
  theme_minimal()
``` 


## Normality of Residuals
- This does not look like a normal distribution, which violates the assumptions of the linear mixed model. We could try to log-transform the response times to see if that helps with normality, but this can make interpretation more difficult and may not fully address the issue of skewness [@lo2015transform]. 
- We will address this in the next session using Bayesian modeling with `brms` and ex-Gaussian distributions.

## Summary of Session 1

- We introduced a dataset from a flanker experiment that investigates parafoveal processing of articles with concordant or discordant grammatical gender in Spanish.
- We applied standard preprocessing steps to clean the data and defined custom contrasts to test our specific hypotheses about the effects of the flanker conditions.
- We were unable to fit a maximal random effects structure due to convergence issues, which is a common problem in `lme4`. We simplified the random effects structure while retaining the most important random slopes related to our primary hypotheses.

## Summary of Session 1 (continued)

- The fitted model revealed significant facilitation effects for the concordant flankers and significant interference effects for the discordant flankers, as well as a main effect of stimulus type.
- There were issues with our fitted model: we were not able to fit the maximal random effects structure, and the residuals of the model were not normally distributed due to the skewed distribution of response times. We will address these issues in the next session using Bayesian modeling with `brms` and ex-Gaussian distributions.

## Homework for Session 2

- Find a data set from your own research or from the literature that you would like to analyze using linear mixed models. It should ideally have a similar structure to the flanker dataset (i.e., repeated measures with crossed random effects of subjects and items).
- Follow the same preprocessing steps we applied to the flanker data (e.g., data cleaning, applying contrast coding) and fit a linear mixed model using `lme4`. Try to include a maximal random effects structure and see if it converges. If not, try simplifying the random effects structure while retaining the most important random slopes.

## References

::: {#refs}
:::

