---
title: "Session 2: Bayesian Mixed-Effects Models with brms"
subtitle: "Data Analysis Workshop 2026"
author: "Dr. Bernhard Angele"
format:
  revealjs:
    incremental: true
---

## Session roadmap (3h)

- **Part 1 (35 min):** frequentist-to-Bayesian translation on the same dataset
- **Part 2 (55 min):** prior specification and family selection
- **Part 3 (60 min):** fitting, diagnostics, and model comparison
- **Part 4 (30 min):** posterior interpretation and reporting

## Continuity from Session 1

Same dataset and question as Session 1:

- Data: `analysis-workshop/data/Hindi_new.csv`
- Outcome: first-run dwell time
- Predictors: `writing_system`, `frequency`, and their interaction
- Random effects: participants and items

## Why Bayesian for this case?

- Right-skewed RT data makes Gaussian assumptions fragile
- Complex random-effects structures may be weakly identified
- We want direct probability statements (e.g., probability effect < 0)
- Priors let us regularize implausible parameter values

## Bayesian workflow for this dataset

1. Reuse cleaned dataset from Session 1
2. Define priors from domain scale
3. Fit competing likelihood families
4. Diagnose chains and geometry
5. Compare predictive performance
6. Summarize posterior inferences for hypotheses

## Step 1: load packages and data

```{r}
#| eval: false
library(tidyverse)
library(brms)
library(tidybayes)
library(bayesplot)
library(loo)

hindi <- read_csv("analysis-workshop/data/Hindi_new.csv") %>%
  mutate(
    subject = factor(RECORDING_SESSION_LABEL),
    item = factor(stim_num),
    writing_system = if_else(cond_name %in% c("RL", "RH"), "Romance", "Traditional"),
    frequency = if_else(cond_name %in% c("RH", "TH"), "High", "Low"),
    writing_system = factor(writing_system),
    frequency = factor(frequency)
  ) %>%
  filter(practice == "No", IA_SKIP == 0) %>%
  mutate(
    frd = if_else(
      IA_FIRST_RUN_DWELL_TIME >= 80 & IA_FIRST_RUN_DWELL_TIME <= 2000,
      IA_FIRST_RUN_DWELL_TIME,
      NA_real_
    )
  )
```

## Decision checkpoint A: response family candidates

- Candidate 1: `gaussian()` on `log(frd)`
- Candidate 2: `shifted_lognormal()` on raw `frd`
- Candidate 3: `exgaussian()` on raw `frd`

**Decision strategy:** keep all three initially, then compare posterior predictive fit and LOO.

## Step 2: scale and coding decisions

```{r}
#| eval: false
hindi <- hindi %>%
  mutate(log_frd = log(frd))

contrasts(hindi$writing_system) <- contr.sum(2)
contrasts(hindi$frequency) <- contr.sum(2)
```

- Sum coding keeps parameter interpretation aligned with Session 1

## Step 3: priors based on plausible RT scale

```{r}
#| eval: false
priors_gaussian <- c(
  prior(normal(6, 0.6), class = "Intercept"),  # log-ms scale
  prior(normal(0, 0.2), class = "b"),
  prior(exponential(1), class = "sd"),
  prior(lkj(2), class = "cor"),
  prior(exponential(1), class = "sigma")
)

priors_rt <- c(
  prior(normal(500, 250), class = "Intercept"),
  prior(normal(0, 80), class = "b"),
  prior(exponential(1), class = "sd"),
  prior(lkj(2), class = "cor")
)
```

## Prior predictive check before fitting

```{r}
#| eval: false
fit_prior <- brm(
  log_frd ~ writing_system * frequency + (1 + frequency | subject) + (1 + frequency | item),
  data = hindi,
  family = gaussian(),
  prior = priors_gaussian,
  sample_prior = "only",
  chains = 4, iter = 2000, warmup = 1000
)

pp_check(fit_prior)
```

- **Decision:** if prior predictive mass is unrealistic, tighten priors before posterior fitting

## Step 4: fit Gaussian model on log RT

```{r}
#| eval: false
fit_gaussian <- brm(
  log_frd ~ writing_system * frequency + (1 + frequency | subject) + (1 + frequency | item),
  data = hindi,
  family = gaussian(),
  prior = priors_gaussian,
  chains = 4, iter = 4000, warmup = 2000,
  cores = 4,
  control = list(adapt_delta = 0.95)
)
```

## Step 5: fit shifted lognormal on raw RT

```{r}
#| eval: false
fit_shifted <- brm(
  frd ~ writing_system * frequency + (1 + frequency | subject) + (1 + frequency | item),
  data = hindi,
  family = shifted_lognormal(),
  prior = priors_rt,
  chains = 4, iter = 4000, warmup = 2000,
  cores = 4,
  control = list(adapt_delta = 0.98, max_treedepth = 12)
)
```

## Optional: ex-Gaussian sensitivity model

```{r}
#| eval: false
fit_exgaussian <- brm(
  frd ~ writing_system * frequency + (1 + frequency | subject) + (1 + frequency | item),
  data = hindi,
  family = exgaussian(),
  prior = priors_rt,
  chains = 4, iter = 4000, warmup = 2000,
  cores = 4,
  control = list(adapt_delta = 0.98)
)
```

## Diagnostics checkpoint B: sampler quality

```{r}
#| eval: false
summary(fit_shifted)
plot(fit_shifted)
mcmc_rhat(rhat(fit_shifted))
mcmc_neff(neff_ratio(fit_shifted))
```

Minimum targets:

- $\hat{R} < 1.01$
- high effective sample size
- zero divergences after warmup

## Diagnostics checkpoint C: predictive adequacy

```{r}
#| eval: false
pp_check(fit_gaussian, ndraws = 100)
pp_check(fit_shifted, ndraws = 100)
pp_check(fit_exgaussian, ndraws = 100)
```

- **Decision:** prefer the family that captures skew and tails with fewer systematic mismatches

## Model comparison for final family choice

```{r}
#| eval: false
loo_compare(
  loo(fit_gaussian),
  loo(fit_shifted),
  loo(fit_exgaussian)
)
```

- **Decision rule:** if one model has clearly better expected out-of-sample fit and good diagnostics, adopt it as primary

## Posterior inference for hypotheses

```{r}
#| eval: false
post <- spread_draws(fit_shifted, b_writing_system1, b_frequency1, `b_writing_system1:frequency1`)

mean(post$b_frequency1 < 0)               # P(frequency effect is faster)
mean(post$`b_writing_system1:frequency1` != 0)
```

- Report directional probabilities and 95% credible intervals

## Marginal effects for interpretation

```{r}
#| eval: false
conditional_effects(fit_shifted, effects = "writing_system:frequency")
```

- Use model-based estimated means instead of raw means for final interpretation

## Example decision log (Bayesian)

1. Reused Session 1 cleaning criteria for comparability
2. Tested Gaussian(log RT), shifted lognormal, and ex-Gaussian likelihoods
3. Ran prior predictive checks and adjusted priors to plausible RT range
4. Increased `adapt_delta` when necessary to eliminate divergences
5. Selected final model by combining diagnostics + LOO + theoretical plausibility

## Reporting template (Bayesian)

- **Model:** formula, family, priors, software/version
- **Sampling:** chains, iterations, warmup, control settings
- **Diagnostics:** $\hat{R}$, ESS, divergences, posterior predictive checks
- **Comparison:** LOO/WAIC and rationale for final model
- **Results:** posterior means/medians, credible intervals, directional probabilities

## Final autonomous assignment (2h independent work)

- Choose one dependent variable from your own project
- Write a short decision log (cleaning + model + diagnostics)
- Fit one frequentist and one Bayesian model for the same question
- Submit a 1-page results-style report with justified model choice

## Closing clinic

- Translate your dissertation question into an explicit multilevel model
- Identify one likely diagnostic failure point in advance
- Define decision criteria before fitting final confirmatory model
