---
title: "Session 2: Bayesian Mixed-Effects Models with brms"
subtitle: "Data Analysis Workshop 2026"
author: "Nebrija Workshop"
format:
  revealjs:
    incremental: true
---

## Audience and Goals

- PhD students in cognitive and experimental psychology
- Psycholinguistics and education research contexts
- Goal: practical Bayesian mixed modeling for complex designs

## 1) Bayesian Paradigm Shift

### Escaping Singular Fit Failures

- Bayesian models estimate uncertainty even near zero random-slope variance
- Regularizing priors prevent pathological boundary behavior

### Philosophy of Priors

- Move beyond flat defaults for complex multilevel designs
- Use weakly informative priors on fixed effects and group-level SDs

```{r}
#| eval: false
priors <- c(
  prior(normal(0, 0.5), class = "b"),
  prior(exponential(1), class = "sd")
)
```

## 2) Practical Lab: Custom Response Distributions

### The Reaction-Time Dilemma

- Avoid arbitrary outlier removal as the primary strategy
- Avoid reflexive log-transform-only workflows

### Alternative Likelihoods

```{r}
#| eval: false
fit_shifted <- brm(
  rt ~ condition + (condition | p | subject) + (condition | i | item),
  data = df,
  family = shifted_lognormal(),
  prior = priors
)

fit_exgaussian <- brm(
  rt ~ condition + (condition | p | subject) + (condition | i | item),
  data = df,
  family = exgaussian(),
  prior = priors
)
```

### Ordinal Judgments

```{r}
#| eval: false
fit_ordinal <- brm(
  rating ~ condition + (condition | subject) + (condition | item),
  data = df,
  family = cumulative(link = "logit")
)
```

## 3) MCMC Assessment and Convergence

- Target 2026 standard: $\\hat{R} < 1.01$
- Check ESS and divergent transitions
- Use trace/density diagnostics and posterior predictive checks

```{r}
#| eval: false
summary(fit_shifted)
pp_check(fit_shifted)
```

## 4) Writing Up Bayesian Mixed Models

- Report posterior means/medians and credible intervals
- Add directional probabilities, e.g., $P(b > 0)$
- Optionally include Bayes Factors where justified

```{r}
#| eval: false
library(tidybayes)
library(ggplot2)

posterior <- spread_draws(fit_shifted, b_condition)
ggplot(posterior, aes(x = b_condition)) +
  geom_density(fill = "steelblue", alpha = 0.4)
```

## Pre-Workshop Assignment and Clinic

- Bring a personal dataset or replication target paper
- Final 30-minute clinic: map dissertation designs to `lme4` or `brms` formulas
