---
title: "Session 1: Frequentist Mixed-Effects Models with lme4"
subtitle: "Data Analysis Workshop 2026"
author: "Dr. Bernhard Angele"
format:
  revealjs:
    incremental: true
---

## Session roadmap (3h)

- **Part 1 (35 min):** dataset, design, and research question
- **Part 2 (55 min):** preprocessing decisions and exploratory plots
- **Part 3 (60 min):** model building, diagnostics, and model revision
- **Part 4 (30 min):** interpretation, reporting, and transition to Bayesian analysis

## This session uses one concrete dataset

- Dataset: `analysis-workshop/data/Hindi_new.csv`
- Paradigm: Hindi eye-tracking during reading
- Design: 2 × 2 factorial (`writing_system` × `frequency`)
- Outcome focus for Session 1: first-run dwell time (`IA_FIRST_RUN_DWELL_TIME`)

## Research question for the walkthrough

Do writing system and lexical frequency influence first-pass reading time, while accounting for participant and item variability?

- H1: High-frequency targets are read faster
- H2: Writing system influences processing time
- H3: Frequency effect may differ by writing system

## Why mixed-effects models here?

- Each participant contributes many trials
- Each item appears across conditions
- We need inference that generalizes to new participants and new items
- This requires crossed random effects, not only fixed-effects ANOVA

## Analysis workflow we will follow

1. Load and inspect raw data
2. Recode factors from `cond_name`
3. Make explicit cleaning decisions
4. Visualize distributions and condition means
5. Fit maximal design-justified model
6. Diagnose warnings and simplify if needed
7. Report effects and model decisions transparently

## Step 1: load packages and data

```{r}
#| eval: false
library(tidyverse)
library(lme4)
library(lmerTest)
library(emmeans)
library(performance)

hindi_raw <- read_csv("analysis-workshop/data/Hindi_new.csv")
```

## Step 2: inspect raw structure

```{r}
#| eval: false
glimpse(hindi_raw)
summary(hindi_raw$IA_FIRST_RUN_DWELL_TIME)
table(hindi_raw$cond_name, useNA = "ifany")
```

- Check if all four conditions are present (`RL`, `RH`, `TL`, `TH`)
- Check missingness and implausible values before modeling

## Step 3: recode experimental factors

```{r}
#| eval: false
hindi <- hindi_raw %>%
  mutate(
    subject = factor(RECORDING_SESSION_LABEL),
    item = factor(stim_num),
    writing_system = case_when(
      cond_name %in% c("RL", "RH") ~ "Romance",
      cond_name %in% c("TL", "TH") ~ "Traditional"
    ),
    frequency = case_when(
      cond_name %in% c("RH", "TH") ~ "High",
      cond_name %in% c("RL", "TL") ~ "Low"
    ),
    writing_system = factor(writing_system),
    frequency = factor(frequency)
  )
```

## Decision checkpoint A: which rows are valid trials?

- **Decision:** exclude practice trials
- **Decision:** keep only fixated targets (`IA_SKIP == 0`)
- **Reason:** practice and skipped targets are not valid for first-pass duration modeling

```{r}
#| eval: false
hindi <- hindi %>%
  filter(practice == "No", IA_SKIP == 0)
```

## Decision checkpoint B: plausible RT window

- Eye-tracking durations can include sensor/mapping artifacts
- **Decision:** keep first-run dwell times between 80 and 2000 ms
- **Reason:** remove implausibly short/long values while keeping realistic cognitive variability

```{r}
#| eval: false
hindi <- hindi %>%
  mutate(
    frd = if_else(
      IA_FIRST_RUN_DWELL_TIME >= 80 & IA_FIRST_RUN_DWELL_TIME <= 2000,
      IA_FIRST_RUN_DWELL_TIME,
      NA_real_
    )
  )
```

## Audit the impact of cleaning decisions

```{r}
#| eval: false
hindi %>%
  summarise(
    n_rows = n(),
    missing_frd = sum(is.na(frd)),
    missing_rate = mean(is.na(frd))
  )
```

- Always report how many observations were excluded and why

## Descriptive statistics by condition

```{r}
#| eval: false
hindi %>%
  group_by(writing_system, frequency) %>%
  summarise(
    n = sum(!is.na(frd)),
    mean_frd = mean(frd, na.rm = TRUE),
    sd_frd = sd(frd, na.rm = TRUE)
  )
```

## Visual check: condition distributions

```{r}
#| eval: false
ggplot(hindi, aes(x = writing_system, y = frd, fill = frequency)) +
  geom_boxplot(outlier.shape = NA, alpha = .5) +
  geom_jitter(width = .15, alpha = .15, size = .8) +
  labs(
    x = "Writing system",
    y = "First-run dwell time (ms)",
    fill = "Frequency"
  ) +
  theme_minimal()
```

## Modeling decision C: transform or not?

- Raw dwell times are typically right-skewed
- We compare:
  - `frd` model on raw scale
  - `log_frd` model on log scale
- **Decision rule:** prefer log model if residual diagnostics improve clearly

```{r}
#| eval: false
hindi <- hindi %>% mutate(log_frd = log(frd))
```

## Contrast coding decision

- With interactions, default treatment coding can obscure main effects
- **Decision:** use sum coding to interpret main effects as grand-mean deviations

```{r}
#| eval: false
contrasts(hindi$writing_system) <- contr.sum(2)
contrasts(hindi$frequency) <- contr.sum(2)
```

## Maximal design-justified model

```{r}
#| eval: false
m_max <- lmer(
  log_frd ~ writing_system * frequency +
    (1 + writing_system * frequency | subject) +
    (1 + writing_system * frequency | item),
  data = hindi,
  REML = FALSE,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)
```

## Typical issue: singular fit

- In small/medium samples, full random-correlation structures may be overparameterized
- **Decision:** remove random-effect correlations first (`||`) before dropping slopes

```{r}
#| eval: false
m_uncor <- lmer(
  log_frd ~ writing_system * frequency +
    (1 + writing_system * frequency || subject) +
    (1 + writing_system * frequency || item),
  data = hindi,
  REML = FALSE
)
```

## If singularity persists: principled simplification

- Keep slopes directly tied to confirmatory hypotheses
- Remove weakest/least identifiable slope terms one step at a time
- Save every model variant and diagnostic result

```{r}
#| eval: false
m_final <- lmer(
  log_frd ~ writing_system * frequency +
    (1 + frequency || subject) +
    (1 + frequency || item),
  data = hindi,
  REML = FALSE
)
```

## Diagnostics checklist for final model

```{r}
#| eval: false
isSingular(m_final, tol = 1e-4)
check_model(m_final)
anova(m_final)
```

- Convergence warning absent
- Singularity status acceptable
- Residual patterns reasonable for interpretation

## Effect estimation and planned comparisons

```{r}
#| eval: false
summary(m_final)
emmeans(m_final, ~ writing_system * frequency)
pairs(emmeans(m_final, ~ frequency | writing_system))
```

## Example decision log (report this in Methods)

1. Excluded practice trials and skipped targets (`IA_SKIP == 1`)
2. Winsorized by exclusion window (80–2000 ms on first-run dwell)
3. Used log-transformed dependent variable after skew inspection
4. Started maximal random structure, simplified by removing correlations first
5. Retained interaction term because theoretically predicted

## Reporting template (frequentist)

- **Data processing:** exclusion criteria and retained sample size
- **Model formula:** fixed effects and random-effects structure
- **Diagnostics:** convergence/singularity and residual checks
- **Results:** fixed effects (estimate, SE, t, p, CI) + estimated marginal means
- **Transparency:** what was tried and why final model was chosen

## Bridge to Session 2 (Bayesian)

- We will keep the **same dataset and research question**
- We will revisit key decisions:
  - likelihood family
  - random-effects regularization
  - interpretation of uncertainty
- This makes frequentist vs Bayesian choices directly comparable

## Key references

- Barr, D. J., Levy, R., Scheepers, C., & Tily, H. J. (2013). Random effects structure for confirmatory hypothesis testing: Keep it maximal. *Journal of Memory and Language, 68*(3), 255-278.
- Matuschek, H., Kliegl, R., Vasishth, S., Baayen, H., & Bates, D. (2017). Balancing Type I error and power in linear mixed models. *Journal of Memory and Language, 94*, 305-315.
- Bates, D., Mächler, M., Bolker, B., & Walker, S. (2015). Fitting linear mixed-effects models using lme4. *Journal of Statistical Software, 67*(1), 1-48.
