---
title: "Session 2: Bayesian Mixed-Effects Models with brms"
subtitle: "Data Analysis Workshop 2026"
author: "Dr. Bernhard Angele"
format:
  revealjs:
    incremental: true
---

## Session roadmap (3h)

- **Part 1 (40 min):** Bayesian multilevel fundamentals and prior specification
- **Part 2 (60 min):** fitting `brms` models for Gaussian and non-Gaussian outcomes
- **Part 3 (50 min):** convergence diagnostics, posterior checks, and model comparison
- **Part 4 (30 min):** scientific reporting and final assignment guidance

## Session goals

- Translate frequentist mixed models into Bayesian formulations
- Specify regularizing priors and understand their practical role
- Fit and interpret `brms` models for RT and ordinal outcomes
- Produce defensible Bayesian write-ups for peer-reviewed publications

## Why move from frequentist to Bayesian mixed models?

- Frequentist maximal models often encounter convergence/singularity limits
- Bayesian estimation can remain stable with weakly informative regularization
- Posterior inference supports direct probability statements about effects

## Frequentist-to-Bayesian mapping

- Formula structure remains familiar (`lme4`-style syntax)
- Add priors explicitly for fixed effects and variance parameters
- Evaluate posterior quality (not only point-estimate significance)

## Priors: practical defaults for workshop analyses

```{r}
#| eval: false
library(brms)

priors <- c(
  prior(normal(0, 0.5), class = "b"),
  prior(exponential(1), class = "sd"),
  prior(lkj(2), class = "cor")
)
```

- Priors regularize estimation and reduce extreme, weakly identified solutions

## Basic Bayesian mixed model

```{r}
#| eval: false
fit_gaussian <- brm(
  rt ~ condition + (condition | p | subject) + (condition | i | item),
  data = df,
  family = gaussian(),
  prior = priors,
  chains = 4,
  cores = 4,
  iter = 4000
)
```

## Modeling reaction times without over-simplification

- Avoid relying only on aggressive trimming or log transforms
- Use response distributions adapted to RT skewness

```{r}
#| eval: false
fit_shifted <- brm(
  rt ~ condition + (condition | p | subject) + (condition | i | item),
  data = df,
  family = shifted_lognormal(),
  prior = priors
)

fit_exgaussian <- brm(
  rt ~ condition + (condition | p | subject) + (condition | i | item),
  data = df,
  family = exgaussian(),
  prior = priors
)
```

## Ordinal outcomes in cognitive/educational research

```{r}
#| eval: false
fit_ordinal <- brm(
  rating ~ condition + (condition | subject) + (condition | item),
  data = df,
  family = cumulative(link = "logit"),
  prior = priors
)
```

- Useful for Likert judgments and acceptability ratings

## MCMC quality checks (minimum standards)

- $\hat{R} < 1.01$
- Adequate effective sample sizes (bulk and tail ESS)
- No problematic divergence patterns

```{r}
#| eval: false
summary(fit_shifted)
plot(fit_shifted)
pp_check(fit_shifted)
```

## Model adequacy and comparison

- Posterior predictive checks for distributional fit
- Compare plausible families (e.g., shifted lognormal vs ex-Gaussian)
- Use LOO/WAIC or theoretically constrained comparisons where appropriate

```{r}
#| eval: false
loo(fit_shifted, fit_exgaussian)
```

## Interpreting posterior effects

- Report posterior estimates and credible intervals
- Add directional probabilities, e.g., $P(\beta > 0)$
- Use visual posterior summaries to support narrative claims

```{r}
#| eval: false
library(tidybayes)
library(ggplot2)

posterior <- spread_draws(fit_shifted, b_condition)
mean(posterior$b_condition > 0)
```

## Bayes factors (when justified)

- Bayes factors are prior-sensitive and require explicit prior reporting
- If used, fit with `sample_prior = "yes"` and report sensitivity checks

```{r}
#| eval: false
hypothesis(fit_shifted, "b_condition = 0")
```

## Writing Bayesian methods and results

- Describe prior rationale, model family, sampling settings, and diagnostics
- Report posterior summaries and uncertainty clearly
- Connect model choice to research question and data-generating process

## Final autonomous assignment (2h independent work)

- Choose `lme4` or `brms` with your own data (or provided public dataset)
- Produce a maximum 1-page results-style report
- Justify model architecture and interpret key effects

## Closing clinic

- Translate each dissertation project into a concrete analysis plan
- Identify minimum reproducible reporting elements for submission-ready analyses
