---
title: "Session 1: Frequentist Mixed-Effects Models with lme4"
subtitle: "Data Analysis Workshop 2026"
author: "Dr. Bernhard Angele"
format:
  revealjs:
    incremental: true
---

## Session roadmap (3h)

- **Part 1 (45 min):** design-model alignment and random-effects structures
- **Part 2 (45 min):** model specification, contrasts, and interpretation
- **Part 3 (60 min):** convergence/singularity diagnostics and troubleshooting
- **Part 4 (30 min):** reporting, APA-style write-up, and project clinic

## Workshop context

- Audience: doctoral researchers in cognitive psychology, language, and education
- Focus: applied, critical, publication-oriented mixed-effects analysis
- Core software: `lme4`, `lmerTest`, `emmeans`, `ggplot2`

## Learning objectives for this session

- Justify complex crossed random-effects structures for participants and items
- Diagnose and resolve convergence and singular-fit warnings in `lme4`
- Specify defensible contrasts tied to hypotheses
- Draft method/results reporting aligned with journal expectations

## Why mixed-effects models?

- Repeated measures violate independence assumptions
- Participants and stimulus items add crossed random variability
- Ignoring random structure inflates error rates and weakens generalization

## The "keep it maximal" discussion

- Barr et al. (2013): maximal random-effects structures for confirmatory testing
- Matuschek et al. (2017): balance Type I error control and statistical power
- Practical rule: start from design-justified maximal model, simplify transparently only when needed

## Crossed vs nested random effects

- **Crossed:** same participants see multiple items and conditions
- **Nested:** lower-level units appear only inside one higher-level unit
- Misclassification changes uncertainty estimates and inferential validity

## Core lme4 model architecture

```{r}
#| eval: false
library(lme4)

model <- lmer(
  rt ~ condition * trial_z +
    (1 + condition * trial_z | subject) +
    (1 + condition | item),
  data = df
)
```

- Fixed effects capture experimental hypotheses
- Random effects capture participant/item-level heterogeneity

## Contrast coding and interpretation

- Treatment coding can obscure main-effect interpretation in interactions
- Sum/Helmert/custom contrasts align coefficients with planned hypotheses

```{r}
#| eval: false
contrasts(df$condition) <- contr.sum(2)
contrasts(df$group) <- contr.helmert(3)
```

## Data preparation before fitting

- Center/scale continuous predictors
- Inspect missing values and impossible values
- Audit outlier policy and document decisions

```{r}
#| eval: false
df$trial_z <- scale(df$trial, center = TRUE, scale = TRUE)
summary(df)
```

## Frequent convergence issues

- **Non-convergence:** optimizer cannot find stable solution
- **Singular fit:** one or more variance components collapse near zero
- These are not interchangeable warnings and need different responses

## Troubleshooting sequence

1. Check data scaling and model complexity mismatch
2. Increase optimizer budget / change optimizer
3. Remove unjustified correlation parameters with `||`
4. Simplify random slopes stepwise with written justification

```{r}
#| eval: false
model_try <- lmer(
  rt ~ condition * trial_z +
    (condition + trial_z || subject) +
    (condition | item),
  data = df,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)
```

## Comparing random-effects specifications

- Start from theoretically justified maximal architecture
- Compare with reduced variants only when diagnostics require it
- Keep a reproducible model log for reviewer transparency

```{r}
#| eval: false
anova(model_maximal, model_reduced)
```

## Interpreting and reporting frequentist models

- Report fixed effects with estimates, SE, test statistic, and intervals
- Report random-effect structure and convergence outcomes
- Explicitly state simplification decisions and why they were needed

```{r}
#| eval: false
library(emmeans)
library(modelsummary)

modelsummary(model_try, statistic = "conf.int", output = "markdown")
emmeans(model_try, ~ condition)
```

## APA-style reporting template

- **Method:** design, coding scheme, model formula, optimizer settings
- **Results:** fixed effects, uncertainty metrics, random-effects diagnostics
- **Figures:** effect visualizations and model-based contrasts

## Hands-on activity (guided)

- Apply session workflow to one of your own dissertation variables
- Draft a candidate maximal model for participants and items
- Perform one diagnostic cycle and document your decisions

## Preparation for Session 2 (Bayesian)

- Identify one frequentist model from today with unstable/singular behavior
- Bring the same dataset for Bayesian re-fitting in `brms`
- Note where priors and alternative likelihoods could improve inference

## Key references

- Barr et al. (2013)
- Matuschek et al. (2017)
- Bates et al. (2015) `lme4`

## Questions and discussion

- Final Q&A and project clinic (last 30 minutes)
