---
title: "Session 1: Frequentist Mixed-Effects Models with lme4"
subtitle: "Data Analysis Workshop 2026"
author: "Nebrija Workshop"
format:
  revealjs:
    incremental: true
---

## Audience and Goals

- PhD students in cognitive and experimental psychology
- Psycholinguistics and education research contexts
- Goal: defensible frequentist mixed-effects workflows in `lme4`

## 1) Design-Model Alignment

### The "Maximal" Debate

- Barr et al. (2013): strict design-model alignment to protect Type I error
- Matuschek et al. (2017): pragmatic simplification to preserve power
- Practice: pre-register model architecture and simplification logic

### Crossed vs. Nested Factors

- Crossed: participants see multiple items/lists over conditions
- Nested: lower-level units only occur within one higher-level grouping
- Incorrect classification biases uncertainty and inferential claims

## 2) Advanced `lme4` Mechanics & Syntax

### Suppressing Correlations with `||`

- `(Condition || Subject)` fits intercept + slope variances without correlation terms
- Useful when correlation estimates are unstable or weakly identified
- Assumes random intercept/slope correlation can be omitted for estimation stability

### Dummy vs. Contrast Coding

- Treatment coding can make slope variance interpretation fragile
- Use sum/Helmert/custom contrasts for interpretable main effects

```{r}
#| eval: false
contrasts(df$Condition) <- contr.sum(2)
model <- lmer(rt ~ Condition + (Condition || Subject) + (Condition || Item), data = df)
```

## 3) Hands-On Lab: Convergence Crisis Routine

### Diagnostics

- Distinguish singular fit (`boundary (singular) fit`) from non-convergence
- Inspect variance components and gradient warnings

### Troubleshooting Pipeline

1. Scale/center continuous predictors
2. Try optimizer and iteration controls
3. Simplify random effects stepwise and justify each decision

```{r}
#| eval: false
df$z_trial <- scale(df$trial, center = TRUE, scale = TRUE)
model <- lmer(
  rt ~ Condition * z_trial + (Condition + z_trial | Subject) + (Condition | Item),
  data = df,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)
```

## 4) Reporting & Publication Formatting

### Method Section Blueprint

- State fixed effects, random effects, optimizer, and convergence outcomes
- Document any principled simplification from maximal starting structure

### Extracting Tables

```{r}
#| eval: false
library(modelsummary)
modelsummary(model, statistic = "conf.int", output = "markdown")
```

- Alternatives: `sjPlot::tab_model()` for publication-ready tables

## Pre-Workshop Assignment

- Bring a dataset from your own dissertation/project
- Or bring a target paper to replicate
- Final 30-minute clinic: draft exact `lme4`/`brms` formulas for each project
